Life in the canopy: Community trait assessments reveal substantial functional diversity among fern epiphytes

### 2020-01-15 ms/manuscript.Rmd
Update MS
various fixes to output docx correctly

# Set up captions
# - Figures
fig_nums <- captioner(prefix = "Fig.")
fig_nums(name = "traits-used-image", caption = "Examples of traits used in this study.")
fig_nums(name = "climate", caption = "Microclimatic variables.")
fig_nums(name = "pca", caption = "Trait PCA.")
fig_nums(name = "tree", caption = "Traits on tree.")
fig_nums(name = "heatmap", caption = "Variable importance scores.")
fig_nums(name = "cwm-div", caption = "Community-weighted means.")
fig_nums(name = "comm-div", caption = "Community diversity metrics.")

@@

# SI figures

# Make short versions of citation functions
figure <- pryr::partial(fig_nums, display = "cite")
table <- pryr::partial(table_nums, display = "cite")
s_figure <- pryr::partial(s_fig_nums, display = "cite")
s_table <- pryr::partial(s_table_nums, display = "cite")

<!-- Work-around to get pandoc to recognize custom siunitx units -->
\renewcommand{\hour}{h}
\renewcommand{\second}{s}
\renewcommand{\ma}{MA}
\renewcommand{\year}{yr}
\renewcommand{\month}{mo}
\renewcommand{\week}{wk}
\renewcommand{\day}{d}
\renewcommand{\basepair}{bp}

@@

^1^Department of Organismic and Evolutionary Biology and Harvard University Herbaria, Harvard University, Cambridge, MA 02138, U.S.A.

^2^Department of Biology, Colgate University, 13 Oak Drive, Hamilton, New York 13346, U.S.A.

^3^Current affiliation: Department of Botany, National Museum of Natural History, Smithsonian Institution, Washington, D.C., 20013, U.S.A.

Authors for correspondence:  
*Joel H. Nitta*  
*Tel: +1 2026331000*  
*Email: joelnitta@gmail.com*

*Charles C. Davis*  
*Tel: +1 6174960515*  
*Email: cdavis@oeb.harvard.edu*

```{r word-count table}
tibble(
  words = c(
    "Total word count (Introduction, Materials and Methods, Results, Discussion):_n",
    "Summary:_n",
    "Introduction:_n",
    "Materials and Methods:_n",
    "Results:_n",
    "Discussion:_n",
    "Acknowledgements:_n"),
  figs_tabs = c(
    "No. of figures:_7 (all in color)",
    "No. of tables:_5",
    "No. of supporting information files:_8 (Figs. S1-S2; Tables S1-S6)", "", "", "", "")
) %>%
  separate(words, c("words", "word_count"), sep = "_") %>%
  separate(figs_tabs, c("figs", "fig_count"), sep = "_", fill = "right") %>%
  mutate_all(~replace_na(., "")) %>%
  knitr::kable(format = "latex", booktabs = TRUE, escape = TRUE, col.names = rep("", ncol(.))) %>%
  kable_styling(full_width = FALSE) %>%
  column_spec(1, width = "12em") %>%
  column_spec(2, width = "2em") %>%
  column_spec(3, width = "12em") %>%
  column_spec(4, width = "8em")
```

@@

To gain insight into the role of the canopy in promoting or restricting diversity, a functional and community-based approach can be applied to study this question. 
If the canopy is uniformly harsher than the terrestrial environment, thus acting as a 'filter', we would expect functional convergence of epiphytes at the community level, which in turn may result in decreased phylogenetic diversity of epiphytic communities if traits are phylogenetically conserved [@Webb2002; @Cavender-Bares2009]. 
Conversely, the stratification of the canopy may stimulate functional diversification by niche partitioning, leading to increased functional diversity of epiphytic communities. 
Furthermore, the role of the canopy as a promoter or restrictor of diversity may be modulated by overall abiotic conditions, including most notably relative humidity (RH) and temperature. 
For instance, elevation is a strong predictor of fern species richness, with maximal richness occurring at mid-elevations, especially in the tropics [@Kromer2005; @Kessler2011a]. 
Functional diversity is also correlated with species richness [@Schleuter2010]. 
Thus, environmental variation along elevational gradients is expected to impact the functional, phylogenetic, and species diversity of epiphytic and terrestrial fern communities.

@@

### Study Site

Moorea, French Polynesia (17°30'S, 149°50'W) is a small (\SI{135}{\kilo\meter}^2^) tropical oceanic island more than \SI{5000}{\kilo\meter} from the nearest continental landmass.
This distance acts as a strong barrier to dispersal [@Carson1995; @Dassler2001], which combined with the young age of the island (\circa{ }\SI{1.5}{\ma} [@Duncan1976]) has led to a relatively small, yet phylogenetically diverse fern flora (\circa{ }130 \species, 8/11 orders; taxonomy follows @PteridophytePhylogenyGroupI2016; `r s_table("species-list")`).
Ferns of Moorea include representatives of multiple Cretaceous epiphytic radiations (\eg, *Asplenium*, *Elaphoglossum*, Hymenophylloideae, Polypodiaceae, Vittarioideae; @Schuettpelz2009) and their terrestrial relatives, making them suitable for a comparative study of traits related to the evolution of epiphytic growth in ferns.

### Community Survey

A fern community survey was carried out on Moorea comprising 17, \SI{10}{\meter} × \SI{10}{\meter} plots spanning an elevational gradient from \SI{20}{\meter} to \SI{2000}{\meter}, with plots placed \circa every \SI{200}{\meter} [@Nitta2017].
Most plots were located along trails leading to the three main peaks on the island, Mt. Rotui (\SI{899}{\meter}), Mt. Mouaputa (\SI{890}{\meter}), and Mt. Tohiea (\SI{1206}{\meter}).
Each plot was divided into 25, \SI{2}{\meter} × \SI{2}{\meter} subplots.
Presence/absence of all sporophytes in each subplot was scored and summed to produce an abundance ranking for each species per plot from zero (not observed) to 25 (observed in all subplots).
Epiphytic ferns were sampled by hand to \circa{ }\SI{2}{\meter} on tree trunks or confirmed visually from the ground.
Species were categorized as epiphytic or terrestrial based on field observations (see Trait Selection and Measurements below).
The list of species in each plot was split into epiphytic and terrestrial groups and treated separately during data analysis, hereafter referred to as 'epiphytic' and 'terrestrial' communities.
Vouchers were deposited at GH and UC (abbreviations follow Index Herbariorum; @Thiers2019).

### Environmental Survey

@@

### Trait Selection and Measurements

@@

Morphological traits of gametophytes were all qualitative (either categorical or binary; `r table("traits-used")`).
Traits were scored by observing gametophytes that were either collected in the field and identified using DNA barcoding [@Nitta2017] or grown from spores of known species in the lab, taken from the literature, or inferred from taxonomic group (\ie, genus) if documented as a stable taxonomic character.
For comparative analysis, gametophyte morphotypes were aggregated into two categories: either cordate (\ie, heart-shaped) or non-cordate [@Ebihara2013].
Non-cordate morphotypes include elongate (strap or ribbon \sensu{ }@Farrar2008) and filamentous forms (`r figure("traits-used-image")`).
Gametophytes were cultivated on Bold's media [@Bold1957] supplemented with Nitch's micronutrients [@Nitsch1951].

@@

We tested for phylogenetic signal in quantitative traits using Blomberg's \K{ }[@Blomberg2003] and Pagel's \plam{ }[@Pagel1999], two metrics with slightly different interpretations and sensitivity to different evolutionary scenarios [@Munkemuller2012], with the \function{phylosig} function in the \package{phytools} R package [@Revell2012].
\plam is a scaling parameter that ranges from 0 (traits evolving randomly) to 1 (traits evolving along branches according to Brownian motion (BM)).
\K describes the ratio between the amount of observed variance in traits vs. the amount of variance expected under BM: when \K{ }> 1, traits are more conserved than expected under BM; when \K{ }< 1, traits have less phylogenetic signal than expected under BM [@Blomberg2003].
We tested for significance by comparing observed \K against a null distribution of \num{1000} trees with trait values randomized across the tips, and in \plam by conducting a likelihood ratio test comparing the log likelihoods of the observed value of \plam vs. \plam{ }= 0 (no phylogenetic signal).
We tested for phylogenetic signal in qualitative traits using Fritz and Purvis' \D{ }[@Fritz2010].
Values of \D range from 0 under a model of trait evolution by BM, to 1 for random distribution of traits.
We analyzed \D using the \function{phylo.d} function in the \package{caper} R package [@Orme2013], and conducted significance test by comparing observed \D to distributions of simulated values produced under two models: random shuffling of traits on the tree, or simulation of a binary trait under a BM model.
To account for possible circularity in measuring the phylogenetic signal of traits that were scored based on taxonomy, we conducted the same analyses on a dataset excluding species that had been scored this way ('strict' dataset).

@@

Raw values of MPD and MNTD are statistically related to species richness [@Swenson2014]; to obtain metrics that are uncorrelated with species richness, we calculated their standard effect sizes (SES) by comparing observed values to a distribution of \num{999} randomized communities.
We constructed null communities using the 'independent swap' algorithm [@Gotelli2000].
We included all species of ferns from Moorea and the neighboring island of Tahiti with genetic data available [`r Ntip(phy)` \species total; @Nitta2017] for phylogenetic null communities, and all species with trait data available (`r fern_traits %>%pull(species) %>% n_distinct`) for functional null communities.
We used the \function{ses.mpd} and \function{ses.mntd} functions in the R package \package{picante} [@Kembel2010] to calculate phylogenetic and functional diversity metrics.
Hereafter, 'MPD~phy~' and 'MNTD~phy~', and 'MPD~func~' and 'MNTD~func~', refer to the standard effect sizes of MPD and MNTD as measured with phylogenetic and functional trait distances, respectively.

@@

### Environmental Survey

@@

### Principal Components Analysis of Trait Data

@@

### Phylogenetic Signal

@@

### Correlation of Traits with Growth Habit

@@

### Functional and Phylogenetic Diversity

@@

We observed a total of `r traits_with_abun %>% filter(abundance > 0) %>% pull(species) %>% n_distinct` species across all plots, including `r n_epi` epiphytic and `r n_ter` terrestrial spp.
Mean species richness was `r mean_rich %>% filter(habit == "epiphytic") %>% pull(stat)` \species per plot for epiphytes and `r mean_rich %>% filter(habit == "terrestrial") %>% pull(stat)` \species per plot for terrestrial species (not significantly different; two-sided \tval-test, \pval{ }= `r get_mod_result(t_test_print, "ntaxa", "p")`; `r figure("comm-div")`).
Species richness increased with elevation for both epiphytes and terrestrial communities (`r figure("comm-div")`).
Epiphytic communities were more diverse than terrestrial ones as measured by both MPD~func~ and MNTD~func~ (two-sided \tval-test, `r both`\pval{ }`r p_mpd_func_mntd_func`).
Functional diversity increased along the gradient for both terrestrial and epiphytic communities when measured with MPD~func~ (linear model, \rval^2^ = `r get_mod_result(model_summ_print, "mpd.obs.z.func", "r")`, \pval{ }`r get_mod_result(model_summ_print, "mpd.obs.z.func", "p")`), but not MNTD~func~.
Epiphytic communities were more phylogenetically clustered than terrestrial communities overall when measured using MNTD~phy~ (two-sided \tval-test, \pval{ }`r get_mod_result(t_test_print, "mntd.obs.z.phy", "p")`) but not MPD~phy~ (\pval{ }= `r get_mod_result(t_test_print, "mpd.obs.z.phy", "p")`; `r figure("comm-div")`).
No significant trend was detected in phylogenetic community structure with elevation in either terrestrial or epiphytic communities (`r figure("comm-div")`).

@@

Our study reveals a striking duality in the phylogenetic and functional diversity of fern epiphytes. 
On one hand, we find that fern epiphytes are consistently smaller than terrestrial species at both the species and community level (`r figure("cwm-div")`, `r table("pics")`), and that epiphytic communities are phylogenetically clustered (`r figure("comm-div")`).
At the same time, we find that epiphytic fern communities are significantly more diverse functionally compared to terrestrial communities (`r figure("comm-div")`).
Collectively, these results suggest that environmental filtering plays a strong role in limiting plant size, while simultaneously promoting diversification across other trait axes in epiphytic ferns.

@@

Epiphytes are markedly smaller than terrestrial species as measured by stipe and frond length, frond width, and rhizome diameter at both the species and community level (`r figure("cwm-div")`, `r table("pics")`), supporting both hypotheses (convergence in epiphytic traits and reduced functional diversity of epiphytic communities).
These results are corroborated by @Testo2018a, who similarly observed abrupt size reduction associated with epiphytic growth in a globally sampled fern phylogeny. 
What might explain such consistent and convergent reduction in fern sporophyte size when transitioning into the canopy?
An obvious answer is that water limitation in the canopy exerts strong selective pressure to maintain small size. 
In spite of similar and predictable leaf-level economic tradeoffs between angiosperms and ferns [@Karst2007], fern leaves are limited in their hydraulic flexibility as they rely wholly on primary xylem that feeds into laminar tissues with reduced leaf venation.
Stipe length and structure in particular are critical for regulating whole leaf hydraulic conductance in ferns [@Pittermann2013].
Species growing in water-limited habitats like those in the canopy are expected by biomechanical necessity to decrease stipe length to minimize resistance [@WatkinsJr.2010].
Our results support previous studies that identified a prevalence of short stipes among epiphytic ferns [@WatkinsJr.2010; @Creese2011].
However, given that stipe and frond length, frond width, and rhizome diameter were correlated in our study, and similar traits have previously been shown to co-vary [@Arcand2008; @Creese2011], careful experimental manipulations are required to distinguish the role of each trait in facilitating epiphytism.

Another clear and convergent pattern to emerge from our study supporting the hypothesis of convergence is the association between epiphytism and the prevalence of non-cordate gametophyte morphotypes (\eg, elongate or filamentous).
Gametophyte morphotype was previously observed to correlate with life history and habitat in ferns: terrestrial species tend to have short-lived (< \SI{1}{\year}), cordate morphotypes that establish following disturbance and rapidly produce sporophytes, whereas epiphytic species tend to have non-cordate gametophytes that persist over multiple growing seasons and produce sporophytes more slowly [@WatkinsJr.2007a; @Farrar2008].
The complex, three-dimensional structure of non-cordate gametophytes may serve two important functions. 
Complex thallus morphology can slow drying rates, and slow dehydration is correlated with the ability to recover from drought in gametophytes---likely a critical adaptation to drought-prone epiphytic habitats [@WatkinsJr.2007c; @Pittermann2013].
Non-cordate thallus morphology may also promote outcrossing through the formation of long-lived gametophyte banks that can persist in epiphytic but not in more disturbance-prone terrestrial habitats [@WatkinsJr.2007a; @Farrar2008].
Our study is the first to our knowledge to demonstrate that the correlation between non-cordate morphology and epiphytism in ferns is significant across broad phylogenetic scales.

@@

Despite the apparent constraints on size and the recurrence of non-cordate gametophytes exhibited by epiphytes, we observed greater morphological diversity among epiphytes versus terrestrial species for several traits.
These results do not support the hypothesis that fern epiphytes have reduced functional diversity.
For example, although epiphytes demonstrated less dissected fronds at the species level (`r table("pics")`), they exhibited much greater variation in this trait relative to terrestrial ferns at the community level (`r s_figure("cwsd")`).
It is likely that the high diversity of this trait reflects different physiological adaptations to various canopy niches. 
For instance, filmy ferns (Hymenophyllaceae), which depend on desiccation tolerance to survive dry conditions, have extremely thin, highly dissected fronds.
Bird's nest ferns (\eg, *Asplenium nidus* L.), in contrast, avoid desiccation by capturing detritus and water in the bases of their simple, rosette leaves. 
At low-elevation sites on Moorea where these ferns co-occur, filmy ferns frequently grow in more humid portions of trees including near the main trunk, whereas bird's nest ferns often grow higher in the canopy and on more exposed, drought-prone branches (J. Nitta, pers. obs.). 
Similar predictable patterns of niche partitioning at the local scale have been reported in neotropical epiphtyic ferns as well [@Hietz1998; @Tausz2001]. 
This distribution may reflect alternative water-usage strategies aligned to specific microhabitats in the canopy [desiccation tolerance vs. water capture; @Hietz1998]. 
Such hypotheses need to be carefully tested with additional quantitative investigation, including more finely examining microclimatic variation and species composition across the canopy.

@@

We did not find that gemmae production in gametophytes was linked with epiphytism. 
Rather, occurrence of gemmae strongly reflects phylogeny, appearing concentrated in filmy ferns (Hymenophyllaceae) and vittarioid ferns (`r figure("tree")`).
Gemmae are asexual propagules occurring primarily in non-cordate gametophytes that allow gametophytes to persist over multiple growing seasons and attain large population sizes, and have been hypothesized to be associated with harsh environmental conditions [@Farrar2008].
In particular, gemmae production should be useful in an epiphytic context because they may allow gametophytes to increase their boundary layer and hedge against population loss during extreme drying [@Farrar2008].
Gemmae can also function sexually by producing antheridia in the presence of antheridiogen secreted by other gametophytes [@Emigh1977], which may confer a selective advantage in the epiphytic environment by enabling outcrossing [@Dassler2001].
As with our results summarized above, it is possible that the diversity of gemmae and non-gemmae producing epiphytic species reflects undetected niche variation in the canopy.
Interestingly, when we classified gametophytes as 'independent' if they occurred beyond the elevational range of conspecific sporophytes using the data of @Nitta2017, we found that gemmae production is correlated with independent growth (log likelihood dependent model = `r correlated_evo_test_gemmae$dependent.logL %>% round(1) %>% as.numeric`, log likelihood independent model = `r correlated_evo_test_gemmae$independent.logL %>% round(1) %>% as.numeric`, \pval{ }= `r correlated_evo_test_gemmae$P %>% round(3) %>% as.numeric`, Pagel's test of correlated evolution), echoing a pattern observed in Japan of non-cordate gametophytes frequently occurring independently of their conspecific sporophytes [@Ebihara2013].

>>>>

# Set up captions

# - Figures
fig_nums <- captioner(prefix = "Fig.")
fig_nums(name = "traits-used-image", caption = "Examples of traits used in this study.")
fig_nums(name = "climate", caption = "Microclimatic variables.")
fig_nums(name = "pca", caption = "Trait PCA.")
fig_nums(name = "tree", caption = "Traits on tree.")
fig_nums(name = "heatmap", caption = "Variable importance scores.")
fig_nums(name = "cwm-div", caption = "Community-weighted means.")
fig_nums(name = "comm-div", caption = "Community diversity metrics.")

@@

# Make short versions of citation functions
figure <- pryr::partial(fig_nums, display = "cite")
table <- pryr::partial(table_nums, display = "cite")
s_figure <- pryr::partial(s_fig_nums, display = "cite")
s_table <- pryr::partial(s_table_nums, display = "cite")
```

<!-- Work-around to get pandoc to recognize custom siunitx units -->
<!-- see https://github.com/jgm/pandoc/issues/4773 -->
\renewcommand{\hour}{h}
\renewcommand{\second}{s}
\renewcommand{\ma}{MA}
\renewcommand{\year}{yr}
\renewcommand{\month}{mo}
\renewcommand{\week}{wk}
\renewcommand{\day}{d}
\renewcommand{\basepair}{bp}

@@

^1^Department of Organismic and Evolutionary Biology and Harvard University Herbaria, Harvard University, Cambridge, MA 02138, U.S.A.; ^2^Department of Biology, Colgate University, 13 Oak Drive, Hamilton, New York 13346, U.S.A.; ^3^Current address: Department of Botany, National Museum of Natural History, Smithsonian Institution, Washington, D.C., 20013, U.S.A.

Authors for correspondence:  
*Joel H. Nitta*  
*Tel: +1 2026331000*  
*Email: \textnormal{joelnitta@gmail.com}*

*Charles C. Davis*  
*Tel: +1 6174960515*  
*Email: \textnormal{cdavis@oeb.harvard.edu}*

<!-- FIXME: add borders to table in docx -->

```{r word-count table}
tibble(
  words = c(
    "Total word count (Introduction, Materials and Methods, Results, Discussion):_6520",
    "Summary:_198",
    "Introduction:_924",
    "Materials and Methods:_2409",
    "Results:_798",
    "Discussion:_2389",
    "Acknowledgements:_170"),
  figs_tabs = c(
    "No. of figures:_7 (all in color)",
    "No. of tables:_5",
    "No. of supporting information files:_8 (Figs. S1-S2; Tables S1-S6)", "", "", "", "")
) %>%
  separate(words, c("words", "word_count"), sep = "_") %>%
  separate(figs_tabs, c("figs", "fig_count"), sep = "_", fill = "right") %>%
  mutate_all(~replace_na(., "")) %>%
  knitr::kable(format = "latex", escape = TRUE, col.names = NULL) %>%
  kable_styling(full_width = FALSE) %>%
  column_spec(1, width = "12em") %>%
  column_spec(2, width = "2em") %>%
  column_spec(3, width = "12em") %>%
  column_spec(4, width = "8em")
```

@@

To gain insight into the role of the canopy in promoting or restricting diversity, a functional and community-based approach can be applied to study this question. 
If the canopy is uniformly harsher than the terrestrial environment, thus acting as a 'filter', we would expect functional convergence of epiphytes at the community level, which in turn may result in decreased phylogenetic diversity of epiphytic communities if traits are phylogenetically conserved [@Webb2002; @Cavender-Bares2009]. 
Conversely, the stratification of the canopy may stimulate functional diversification by niche partitioning, leading to increased functional diversity of epiphytic communities. 
Furthermore, the role of the canopy as a promoter or restrictor of diversity may be modulated by overall abiotic conditions, including most notably relative humidity (RH) and temperature. 
For instance, elevation is a strong predictor of fern species richness, with maximal richness occurring at mid-elevations, especially in the tropics [@Kromer2005; @Kessler2011a]. 
Functional diversity is also correlated with species richness [@Schleuter2010]. 
Thus, environmental variation along elevational gradients is expected to impact the functional, phylogenetic, and taxonomic diversity of epiphytic and terrestrial fern communities.

@@

### Study site

Moorea, French Polynesia (17°32'S, 149°50'W) is a small (\SI{135}{\kilo\meter}^2^) tropical oceanic island more than \SI{5000}{\kilo\meter} from the nearest continental landmass.
This distance acts as a strong barrier to dispersal [@Carson1995; @Dassler2001], which combined with the young age of the island (\circa{ }\SI{1.5}{\ma} [@Duncan1976]) has led to a relatively small, yet phylogenetically diverse fern flora (\circa{ }130 \species, 8/11 orders; taxonomy follows @PteridophytePhylogenyGroupI2016; `r s_table("species-list")`).
Ferns of Moorea include representatives of multiple Cretaceous epiphytic radiations (\eg, *Asplenium*, *Elaphoglossum*, Hymenophylloideae, Polypodiaceae, Vittarioideae; @Schuettpelz2009) and their terrestrial relatives, making them suitable for a comparative study of traits related to the evolution of epiphytic growth in ferns.

### Community survey

A fern community survey was carried out on Moorea comprising 17, \SI{10}{\meter} × \SI{10}{\meter} plots spanning an elevational gradient from \SI{20}{\meter} to \SI{2000}{\meter}, with plots placed \circa every \SI{200}{\meter} [@Nitta2017].
Most plots were located along trails leading to the three main peaks on the island, Mt. Rotui (\SI{899}{\meter}), Mt. Mouaputa (\SI{890}{\meter}), and Mt. Tohiea (\SI{1206}{\meter}).
Each plot was divided into 25, \SI{2}{\meter} × \SI{2}{\meter} subplots.
Presence/absence of all sporophytes in each subplot was scored and summed to produce an abundance ranking for each species per plot from zero (not observed) to 25 (observed in all subplots).
Epiphytic ferns were sampled by hand to \circa{ }\SI{2}{\meter} on tree trunks or confirmed visually from the ground.
Species were categorized as epiphytic or terrestrial based on field observations (see 'Trait selection and measurements' below).
The list of species in each plot was split into epiphytic and terrestrial groups and treated separately during data analysis, hereafter referred to as 'epiphytic' and 'terrestrial' communities.
Vouchers were deposited at GH and UC (abbreviations follow Index Herbariorum; @Thiers2019).

### Environmental survey

@@

### Trait selection and measurements

@@

Morphological traits of gametophytes were all qualitative (either categorical or binary; `r table("traits-used")`).
Traits were scored by observing gametophytes that were either collected in the field and identified using DNA barcoding [@Nitta2017], grown from spores of known species in the lab, taken from the literature, or inferred from taxonomic group (\ie, genus) if documented as a stable taxonomic character.
For comparative analysis, gametophyte morphotypes were aggregated into two categories: either cordate (\ie, heart-shaped) or non-cordate [@Ebihara2013].
Non-cordate morphotypes include elongate (strap or ribbon \sensu{ }@Farrar2008) and filamentous forms (`r figure("traits-used-image")`).
Gametophytes were cultivated on Bold's media [@Bold1957] supplemented with Nitch's micronutrients [@Nitsch1951].

@@

We tested for phylogenetic signal in quantitative traits using Blomberg's \K{ }[@Blomberg2003] and Pagel's \plam{ }[@Pagel1999], two metrics with slightly different interpretations and sensitivity to different evolutionary scenarios [@Munkemuller2012], with the \function{phylosig} function in the \package{phytools} R package [@Revell2012].
\plam is a scaling parameter that ranges from 0 (traits evolving randomly) to 1 (traits evolving along branches according to Brownian motion (BM)).
\K describes the ratio between the amount of observed variance in traits vs. the amount of variance expected under BM: when \K{ }> 1, traits are more conserved than expected under BM; when \K{ }< 1, traits have less phylogenetic signal than expected under BM [@Blomberg2003].
We tested for significance in \K by comparing observed values of \K against a null distribution of \num{1000} trees with trait values randomized across the tips, and in \plam by conducting a likelihood ratio test comparing the log likelihoods of observed values of \plam vs. \plam{ }= 0 (no phylogenetic signal).
We tested for phylogenetic signal in qualitative traits using Fritz and Purvis' \D{ }[@Fritz2010].
Values of \D range from 0 under a model of trait evolution by BM, to 1 under random distribution of traits.
We analyzed \D using the \function{phylo.d} function in the \package{caper} R package [@Orme2013], and conducted a significance test by comparing observed \D to distributions of simulated values produced under two models: random shuffling of traits on the tree, or simulation of a binary trait under a BM model.
To account for possible circularity in measuring the phylogenetic signal of traits that were scored based on taxonomy, we conducted the same analyses on a dataset excluding species that had been scored this way ('strict' dataset).

@@

Raw values of MPD and MNTD are statistically related to species richness [@Swenson2014]; to obtain metrics that are uncorrelated with species richness, we calculated their standard effect sizes (SES) by comparing observed values to a distribution of \num{999} randomized communities.
We constructed null communities using the 'independent swap' algorithm [@Gotelli2000].
We included all species of ferns from Moorea and the neighboring island of Tahiti with genetic data available [`r Ntip(phy)` \species total; @Nitta2017] for phylogenetic null communities, and all species with trait data available (`r fern_traits %>%pull(species) %>% n_distinct` \species) for functional null communities.
We used the \function{ses.mpd} and \function{ses.mntd} functions in the R package \package{picante} [@Kembel2010] to calculate phylogenetic and functional diversity metrics.
Hereafter, 'MPD~phy~' and 'MNTD~phy~', and 'MPD~func~' and 'MNTD~func~', refer to the standard effect sizes of MPD and MNTD as measured with phylogenetic and functional trait distances, respectively.

@@

### Environmental survey

@@

### Principal components analysis of trait data

@@

### Phylogenetic signal

@@

### Correlation of traits with growth habit

@@

### Functional and phylogenetic diversity

@@

We observed a total of `r traits_with_abun %>% filter(abundance > 0) %>% pull(species) %>% n_distinct` \species across all plots, including `r n_epi` epiphytic and `r n_ter` terrestrial spp.
Mean species richness was `r mean_rich %>% filter(habit == "epiphytic") %>% pull(stat)` \species per plot for epiphytes and `r mean_rich %>% filter(habit == "terrestrial") %>% pull(stat)` \species per plot for terrestrial species (not significantly different; two-sided \tval-test, \pval{ }= `r get_mod_result(t_test_print, "ntaxa", "p")`; `r figure("comm-div")`).
Species richness increased with elevation for both epiphytes and terrestrial communities (`r figure("comm-div")`).
Epiphytic communities were more diverse than terrestrial ones as measured by both MPD~func~ and MNTD~func~ (two-sided \tval-test, `r both`\pval{ }`r p_mpd_func_mntd_func`).
Functional diversity increased along the gradient for both terrestrial and epiphytic communities when measured with MPD~func~ (linear model, \rval^2^ = `r get_mod_result(model_summ_print, "mpd.obs.z.func", "r")`, \pval{ }`r get_mod_result(model_summ_print, "mpd.obs.z.func", "p")`), but not MNTD~func~.
Epiphytic communities were more phylogenetically clustered than terrestrial communities overall when measured using MNTD~phy~ (two-sided \tval-test, \pval{ }`r get_mod_result(t_test_print, "mntd.obs.z.phy", "p")`) but not MPD~phy~ (\pval{ }= `r get_mod_result(t_test_print, "mpd.obs.z.phy", "p")`; `r figure("comm-div")`).
No significant trend was detected in phylogenetic community structure with elevation in either terrestrial or epiphytic communities (`r figure("comm-div")`).

@@

Our study reveals a striking duality in the phylogenetic and functional diversity of fern epiphytes. 
On one hand, we find that fern epiphytes are consistently smaller than terrestrial species at both the species and community level (`r table("pics")`, `r figure("cwm-div")`), and that epiphytic communities are phylogenetically clustered (`r figure("comm-div")`).
At the same time, we find that epiphytic fern communities are significantly more diverse functionally compared to terrestrial communities (`r figure("comm-div")`).
Collectively, these results suggest that environmental filtering plays a strong role in limiting plant size, while simultaneously promoting diversification across other trait axes in epiphytic ferns.

@@

Epiphytes are markedly smaller than terrestrial species as measured by stipe and frond length, frond width, and rhizome diameter at both the species and community level (`r table("pics")`, `r figure("cwm-div")`), supporting both hypotheses (convergence in epiphytic traits and reduced functional diversity of epiphytic communities).
These results are corroborated by @Testo2018a, who similarly observed abrupt size reduction associated with epiphytic growth in a globally sampled fern phylogeny. 
What might explain such consistent and convergent reduction in fern sporophyte size when transitioning into the canopy?
An obvious answer is that water limitation in the canopy exerts strong selective pressure to maintain small size. 
In spite of similar and predictable leaf-level economic tradeoffs between angiosperms and ferns [@Karst2007], fern leaves are limited in their hydraulic flexibility as they rely wholly on primary xylem that feeds into laminar tissues with reduced leaf venation.
Stipe length and structure in particular are critical for regulating whole leaf hydraulic conductance in ferns [@Pittermann2013].
Species growing in water-limited habitats like those in the canopy are expected by biomechanical necessity to decrease stipe length to minimize resistance [@WatkinsJr.2010].
Our results support previous studies that identified a prevalence of short stipes among epiphytic ferns [@WatkinsJr.2010; @Creese2011].
However, given that stipe and frond length, frond width, and rhizome diameter were correlated in our study, and similar traits have previously been shown to co-vary [@Arcand2008; @Creese2011], careful experimental manipulations are required to distinguish the role of each trait in facilitating epiphytism.

Another clear pattern to emerge from our study supporting the hypothesis of convergence is the association between epiphytism and the prevalence of non-cordate (\eg, elongate or filamentous) gametophyte morphotypes.
Gametophyte morphotype was previously observed to correlate with life history and habitat in ferns: terrestrial species tend to have short-lived (< \SI{1}{\year}), cordate morphotypes that establish following disturbance and rapidly produce sporophytes, whereas epiphytic species tend to have non-cordate gametophytes that persist over multiple growing seasons and produce sporophytes more slowly [@WatkinsJr.2007a; @Farrar2008].
The complex, three-dimensional structure of non-cordate gametophytes may serve two important functions. 
Complex thallus morphology can slow drying rates, and slow dehydration is correlated with the ability to recover from drought in gametophytes---likely a critical adaptation to drought-prone epiphytic habitats [@WatkinsJr.2007c; @Pittermann2013].
Non-cordate thallus morphology may also promote outcrossing through the formation of long-lived gametophyte banks that can persist in epiphytic but not in more disturbance-prone terrestrial habitats [@WatkinsJr.2007a; @Farrar2008].
Our study is the first to our knowledge to demonstrate that the correlation between non-cordate morphology and epiphytism in ferns is significant across broad phylogenetic scales.

@@

Despite the apparent constraints on size and the recurrence of non-cordate gametophytes exhibited by epiphytes, we observed greater morphological diversity among epiphytic versus terrestrial communities for several traits.
These results do not support the hypothesis that fern epiphytes have reduced functional diversity.
For example, although epiphytes demonstrated less dissected fronds at the species level (`r table("pics")`), they exhibited much greater variation in this trait relative to terrestrial ferns at the community level (`r s_figure("cwsd")`).
It is likely that the high diversity of this trait reflects different physiological adaptations to various canopy niches. 
For instance, filmy ferns (Hymenophyllaceae), which depend on desiccation tolerance to survive dry conditions, have extremely thin, highly dissected fronds.
Bird's nest ferns (\eg, *Asplenium nidus* L.), in contrast, avoid desiccation by capturing detritus and water in the bases of their simple, rosette leaves. 
At low-elevation sites on Moorea where these ferns co-occur, filmy ferns frequently grow in more humid portions of trees including near the main trunk, whereas bird's nest ferns often grow higher in the canopy and on more exposed, drought-prone branches (J. Nitta, pers. obs.). 
Similar predictable patterns of niche partitioning at the local scale have been reported in neotropical epiphtyic ferns as well [@Hietz1998; @Tausz2001]. 
This distribution may reflect alternative water-usage strategies aligned to specific microhabitats in the canopy [desiccation tolerance vs. water capture; @Hietz1998]. 
Such hypotheses need to be carefully tested with additional quantitative investigation, including more finely examining microclimatic variation and species composition across the canopy.

@@

We did not find that gemmae production in gametophytes was linked with epiphytism. 
Rather, occurrence of gemmae strongly reflects phylogeny, appearing concentrated in filmy ferns (Hymenophyllaceae) and vittarioid ferns (Vittarioideae) (`r figure("tree")`).
Gemmae are asexual propagules occurring primarily in non-cordate gametophytes that allow gametophytes to persist over multiple growing seasons and attain large population sizes, and have been hypothesized to be associated with harsh environmental conditions [@Farrar2008].
In particular, gemmae production should be useful in an epiphytic context because they may allow gametophytes to increase their boundary layer and hedge against population loss during extreme drying [@Farrar2008].
Gemmae can also function sexually by producing antheridia in the presence of antheridiogen secreted by other gametophytes [@Emigh1977], which may confer a selective advantage in the epiphytic environment by enabling outcrossing [@Dassler2001].
As with our results summarized above, it is possible that the diversity of gemmae and non-gemmae producing epiphytic species reflects undetected niche variation in the canopy.
Interestingly, when we classified gametophytes as 'independent' if they occurred beyond the elevational range of conspecific sporophytes using the data of @Nitta2017, we found that gemmae production is correlated with independent growth (log likelihood dependent model = `r correlated_evo_test_gemmae$dependent.logL %>% round(1) %>% as.numeric`, log likelihood independent model = `r correlated_evo_test_gemmae$independent.logL %>% round(1) %>% as.numeric`, \pval{ }= `r correlated_evo_test_gemmae$P %>% round(3) %>% as.numeric`, Pagel's test of correlated evolution), echoing a pattern observed in Japan of non-cordate gametophytes frequently occurring independently of their conspecific sporophytes [@Ebihara2013].
